Bayesian Inference for Species Distribution Modelling with Gaussian Processes

### 2021-04-21 index.Rmd
more

Imagine for a moment you are tossing a fair coin. 
There are many experiments you could perform by tossing a coin, but let us consider our quantity of interest to be the fraction of times our coin lands on a tails. 
It is clear that the number of tails, the outcome of our experiment, is dependent on the eventual realization of some random process.
A random variable $X$ is a variable whose value is dependent on the outcome(s) of a stochastic phenomenon. 
The realized value of $X$ is denoted as $x$.

@@

The r.v. $X$ is is an example of a *discrete random variable*. 
Discrete random variables can only assume discrete values. 
To the contrary, continuous random variables are useful for describing continuous sample spaces.
For example, a continuous random variable may be used to represent the outcome of an experiment measuring blossoming heights of flowers, where the data is a sequence of observations of heights at which different flowers blossomed. 
In this case, the outcome of our blossoming experiment can be any of an infinite number of real values, thus is properly modeled by a continuous random variable.

@@

In the previous example we represented the probability of various outcomes of a coin toss experiment in a tabular format. 
Another way to represent this set of probabilities is as a *cumulative distribution function*.

@@

The cumulative distribution function $F_{X}(x)$ simply represents the probability that a random variable $X$ takes a value less than or equal to $x$ for each possible input value of $x$. 
Figure \@ref(fig:cdfimg) depicts a graphical representation of the CDF for our coin tossing experiment.

@@

Specifically, this implies that $F'_X(x)=f_X(x)$ for all differentiable points of $F_X$. 
In plain English this signifies that the derivative of the CDF is the PDF.

@@

Importantly, while in this case $E(X)$ corresponds well to the "peak" in the PMF, this should not be assumed to be the case unilaterally, as the same expectation would result from any distribution symmetrical about $x=3$.

@@

Conditional probability provides a way to model the probability that an event occurs, given that another event is known to have occured.
Conditional probability, as presented in Definition \@ref(def:condprob), requires only that the event assumed to have occured, i.e., the event we are *conditioning on*, has a nonzero probability of occuring.

@@

Conditional probability asserts that the probability of event A occuring given that event B occurs is equivalent to the probability of both A and B occuring (denoted as $A \cap B$) divided by the probability that B occurs. $P(A|B)$ is not generally equal to $P(B|A)$ 
For example, the probability that I am swimming given that I am in the water is clearly not the same as the probability that I am in the water given that I am swimming.
Example \@ref(exm:condprobe) explains how to use conditional probability to calculate the probability of rolling a 2 on a 6-sided dice, given I know the outcome of the roll is less than 4.

@@

Consider a student, Alice, who was exposed to someone with COVID-19. 
Being a responsible person, Alice decides she should get tested. 
She receives a test and the accompanying information sheet states the test is 85% accurate, meaning that 85% of the time it gives positive results to recipients who are actually positive. The sheet also says that the test yields a false positive 30% of the time, meaning that if Alice is actually negative she will still receive a positive test 30% of the time. 
The following day Alice receives a positive test.
As a student of probability, Alice recognizes that the 85% accuracy statistic only means the conditional probability that she receives a positive test given she is COVID positive ($P(+ test|covid+)$) is 85%.
However, Alice is actually interested in the conditional probability that she is positive given she just tested positive, $P(covid+|test+)$.
As stated in the previous section's discussion of conditional probability, $P(A|B) \neq P(B|A)$.
Bayes' theorem, which follows intuitively from the theorem of conditional probability, provides this answer for Alice.
We can rewrite Definition \@ref(def:condprob) as 
$$P(A \cap B)=P(A|B)P(B).$$
Furthuremore, it can also be stated that
$$P(B \cap A)=P(B|A)P(A)$$
Clearly $P(A \cap B)=P(B \cap A)$, as both terms can be used interchangeably to represent the intersection of two sets A and B.
This equivalency means that we can rewrite these equations as
$$P(B|A)P(A)=P(A|B)P(B)$$
Dividing both sides of this equivalency by $P(B)$ yields Bayes' theorem, as formalized in Definition \@ref(def:bayestheorem) [@Junker2003].

@@

In the previous example, we used Bayes' theorem to evaluate the probability that Alice is COVID positive given she recieved a positive test. 
Bayes' theorem provided a way for Alice to apply her prior beliefs about the likelihood of actually being positive to the data she observed---a positive test. 
*Learning*, or *statistical inference*, asks a generalized version of Alice's question. 
Given some data $X_0,...,X_i \sim D$, how can we infer the distribution $D$ that generated the data $X_0,...,X_i$ [@Wasserman2004 \text{Sec} 6.1]?
Additionally as distributions are functions of parameters $\theta_0, ..., \theta_j$, estimating these parameters given the data is one avenue for inferring $D$.
*Bayesian inference* applies Bayes' theorem to the task of statistical inference in an intuitive manner, which by definition maintains the inherent uncertainty in inference.
Contrary to the frequentist perspective, a Bayesian approach to inference expresses degrees of belief by producing posterior probability distributions for the parameters $\theta_0, ..., \theta_j$ of the distribution [@Wasserman2004 \text{Sec} 11.1]. 
Point estimates and confidence/posterior intervals can then be computed with a post hoc analysis of the posterior distribution.
Furthermore, a Bayesian approach to inference explicitly includes previous domain specific knowledge (or lack thereof) in the learning process.
Performing Bayesian inference to compute a posterior distribution $P(\theta|X)$ for a parameter $\theta$ first requires choosing a *prior probability distribution* $P(\theta)$ with a PDF $f(\theta)$ which expresses our prior beliefs about $\theta$ before observing any data. 
In the previous example, this was Alice's previous belief about the probability of contracting COVID from any given exposure.
Experessing $P(\theta)$ as a distribution allows for the inclusion of the inherent uncertainty of Alice's preconceptions in the modelling schema and thus the computation of the posterior.
Secondly, the likelihod function $l(X|\theta)$ must be specified to express our understanding of the probability of the data given a particular value of $\theta$.
In the case of Alice's inquiry, a binomial likelihood function would be appropriate as it represents the likelihood of observing a given number of successes ($+$tests) in $0 \lor 1$ ($+ \lor -$) trials.
The final step in performing Bayesian inference is repeatedly applying Bayes' theorem to update the probability distribution of $\theta$ for each observation $X_n$ in the data $n \in (0,i)$, resulting in a posterior distribution $f(\theta|X_0,...,X_i)$.
This is done by iteratively evaluating Bayes' theorem, where the posterior for the $n-1^{th}$ iteration is the prior for the $n^{th}$ iteration.
Bayesian inference relies on Bayes' theorem for continuous variables, which utilizes density functions as presented in formally Definition \@ref(def:bayesprob).

@@

While Bayesian inference is an appealing schema, learning the parameters of real-world models using this technique requires the use of efficient algorithms to be computationally feasible.
As previously mentioned, depending on the system we may be interested in evaulating an arbitrary number parameters $\theta_0, ..., \theta_j$ given the data, and each parameter may potentially take on an infinite number of values.
Even in the case where we only consider a small number of possible values $a$ for every $\theta_i \in (0,j)$, the evaluation of the posterior to perform a single update rapidly becomes a computationally intractible task.
This is evidenced by the runtime of the *grid approximation* or *direct discrete approximation*, which is proportional to $O(a^{j})$, where $a$ controls the granularity of the approximation [@Gelman2013 \text{Sec} 10.1].
In order to perform Bayesian inference on complex systems, fast algorithms are necessary to circumnavigate this curse of dimensionality.

@@

When performing Bayesian inference, fast algorithms are necessary for updating the probability distributions of parameters of interest in a computationally feasible manner.
In modelling scenarios with multiple parameters forming a multidimensional parameter space, evaluating parameters at even a relatively small number of possible values rapidly becomes intractible, as a product of the curse of dimensionality.
This problem is further exaggerated in the evaluation of hierarchical models, which are of particular interest in many application domains such as biology, economics, chemistry, and physics.
Here we discuss Markov Chain Monte Carlo (MCMC) based sampling algorithms, which allow for efficiently sampling from approximations of probability distributions.
Markov Chain Monte Carlo (MCMC) provides a methodology for sampling from the posterior curve to perform a Bayesian update. 
Simply put, if we make some sequence of draws $\theta_1,...,\theta_j$ from the posterior $p(\theta|X)$, then plotting $\theta_1,...,\theta_j$ as a histogram will provide an approximation of $p(\theta|X)$ [@Wasserman2004 \text{Sec} 11.4].
Due to the law of large numbers, this approximation can be used in lieu of the posterior in post-hoc analysis.
We first provide a theoretical background into Monte Carlo approximation and Markov Chains, followed by a discussion of *Metropolis-Hastings*, an algorithm for performing MCMC.

@@

Monte Carlo approximation provides a convient method for estimating quantities of interest.
This schema will allow for drawing samples from arbitrarily complex probability distributions.
The basic example is Monte Carlo *integration* [@Wasserman2004 \text{Sec} 24.2]. 
If we want to evaulate an integral for some function $f(x)$, where 
$$I=\int_{a}^{b}f(x)dx \textrm{ ,}$$
we can approximate $I$ using Monte Carlo approximation. $f(x)$ can be alternatively expressed as two functions, $h(x)$ and $w(x)$, where $h(x)=\frac{1}{b-a}$ and $w(x)=f(x)(b-a)$ as follows.
$$I=\int_{a}^{b}f(x)dx=\int_{a}^{b}w(x)h(x)dx$$
Conveniently, $h$ is the pdf of a uniform r.v. $X$ over $[a,b]$, which means that $I$ can be rewritten in terms of expectation as
$$I=E_{f}(w(X)) \textrm{ .}$$
In conjunction with the law of large numbers, this means that if we generate a sequence of random variables from a uniform distribution $X_{0},...,X_{N}\sim unif(a,b)$ then the standard Monte Carlo integration method asserts

@@

```{example montecarl}
Suppose we have two binomially distributed r.vs---$X \sim Binom(n,p_1)$, $Y \sim Binom(m,p_2)$---and want to evaluate $\delta = p_2 - p_1$. 
This problem can be analyzed using Bayesian inference on a beta-binomial model, but would require evaluating complex integrals over the joint posterior of $X$ and $Y$ to obtain the density.
Monte Carlo Integration allows for estimating an integral using repeated sampling from the two independent posterior densities---$P_1$ and $P_2$--- and plotting samples using a histogram provides a qualitative estimate for the parameter of interest, $\delta$ [@Wasserman2004].
In simplest terms the beta prior distribution $Beta(\alpha, \beta)$ (the proper conjugate prior for the binomial likelihood) could be a flat prior $\alpha = \beta = 1$, on the parameters $p_1$ and $p_2$ [@Bloom2014].
$f(p_1,p_2|X,Y)=f(p_1|X)f(p_2|Y)$, because the two posteriors $f(p_1|X)$ and $f(p_2|Y)$ are conditionally independent given the data.
Thus the resultant posteriors are beta distributions, as follows [@Bloom2014].
$$D_1 = Beta(X+1,n-X+1)$$
$$D_2 = Beta(Y+1,m-Y+1)$$
Monte Carlo integration simulates the quantity $\delta$ by drawing $(P_1^{(1)},P_2^{(2)}),...,(P_1^{(N)},P_2^{(N)})$ from the independent beta posteriors $D_1$ and $D_2$ for $i=1,...,N$ assuming N is sufficiently large, every $\delta^{(i)}=P_2^{(i)} - P_1^{(i)}$.
Accordingly, $\delta$ is approximated using the previously established equivalency, $$\delta \approx \frac{1}{N} \sum_i^N \delta^{(i)} = \frac{1}{N} \sum_i^N (P_2^{(i)} - P_1^{(i)})$$ [@Wasserman2004].
To provide a concrete example, if the number of trials is 20, $n=m=20$, $X = 11$, and $Y = 4$ Sampling 10000 times from the in the previously described manner yields the histogram presented in Figure \@ref(fig:mon) 
```
```{r mon, fig.height=3, fig.align='center', echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Histogram displaying displaying posterior of delta from Monte Carlo integration"}
p1 <- rbeta(10000, 12, 10)
p2 <- rbeta(10000, 5, 17)
delta <- p2 - p1
delta_df <- data.frame(delta = delta)
hist <- ggplot(delta_df, aes(x=delta)) + 
  geom_histogram(colour="black", fill="#856cb0") +
  theme_mdsr() +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.title = element_blank())
  #theme(text = element_text(size=12, 
  #                          family = "CMU Serif",
   #                         face = "plain",
   #                         colour = "black",
    #                        hjust = 0.5,
     #                       vjust = 0.5,
      #                      angle = 0,
       #                     lineheight = 0.9))

@@

*Metropolis-Hastings MCMC* is a standard algorithm for sampling from the posterior distribution to perform a Bayesian update. 
While it is not always the fastest in practice and requires manual tuning to work effectively, Metropolis-Hastings provides a foundation for more complex algorithms such as *Hamiltonian Monte Carlo* [@Brooks2011] with *No-U-Turn sampling* [@Homan2014]. 
To reiterate, the purpose of sampling in Bayesian inference is to draw from the posterior density with a goal of estimating $P(\theta|data)$ for parameters $\theta$. 
The Bayesian posterior $P(\theta|data)$, for the purposes of MCMC, is also referred to as the *target distribution*.
The basic idea of Metropolis-Hastings, to return to the previously mentioned example of approximating $I=\int h(x)f(x)dx$, is to construct a Markov chain $X_0,...,$ with a stationary distribution of $h$.
Assuming the sequence has sufficiently many observations, the law of large numbers applies and the following convergence holds.
$$\frac{1}{N}\sum_{i=0}^{N}w(X_{i})\xrightarrow{P} E(w(X))=I$$
The Metropolis-Hastings algorithm provides a method for sampling from the *stationary distribution* of a Markov chain designed in such a way that this distribution approximates the posterior distribution.
Initially, the probability of realizing a state as part of the sampled sequence is skewed, but importantly, over time the sequence will converge.
Critically, the law of large numbers guarantees that with sufficiently many iterations of Metropolis-Hastings, the generated sequence of states will converge on this stationary distribution.
Essentially, given some sequence of states $X_{0},X_{1},...,X_{i}$ from a Markov-Chain, where $X_{0}$ is chosen arbitrarily, an iteration of Metropolis-Hastings produces the next state to include in the sequence. 
As presented formally in Definition \@ref(def:methas), this is achieved by generating a *proposal* for $X_{i+1}$ from the *proposal distribution* $q(y|X_{i})$ and then accepting the proposal with some probability dependent on the relative, target probabilities of the current state $X_{i}$ and the proposal [@Wasserman2004].
```{definition, methas, name="Metropolis-Hastings MCMC"}
*$X_{i+1}$ is generated given $X_{0},X_{1},...,X_{i}$ in the following manner:*
\begin{enumerate}
  \item Sample a proposal $Y$ from the proposal distribution $Y \sim q(y|X_{i}) \textrm{ .}$
  \item Evaulate the ratio of probabilities in the stationary distribution $h(x)$ for $r(X_{i},Y)$ where $$ r(x,y)=min \{\frac{f(y)}{f(x)} \frac{q(x|y)}{q(y|x)} ,1 \} \textrm{ .}$$
  \item Compute the next state $X_{i+1}$, where $$X_{i+1}=\begin{cases}Y & \textrm{with probability }r \\ X_{i} & \textrm{with probability }1-r \end{cases} \textrm{ .}$$
\end{enumerate}
```

While Metropolis-Hastings may seem appealing because it is both memoryless and able to approximate very complex distributions, its downfall lies in the manual tuning of the *step size* parameter necessary to achieve convergence of the stationary distribution to the target distribution in a reasonable number of iterations. 
Recall that Metropolis-Hastings generates proposals by sampling from some distribution.
In the special case of *random walk Metropolis-Hastings*, this is a normal distribution centered about $X_{i}$.
Here, the standard deviation of the normal distribution, known as the step size, dictates the relative distance in the sample space between $X_{i}$ and the generated proposal.
If the step size is too low, the algorithm will make very small steps and may miss key features of the target distribution, causing the stationary distribution to require a much larger number of iterations to converge on the target distribution.
On the other hand if the step size is too large, proposals will be overwhelmingly generated from the low-probability tails of the distribution, again resulting in a lack of convergence and poorly representative samples from the stationary distribution.
While some methods have been proposed for automatically tuning the step size parameter [@Graves2011], iterating on complex, Bayesian models to tune a parameter is not particularly efficient nor computationally feasible in many cases. 
For this reason, other approximation algorithms are used in practical implementations of Bayesian inference.
One such algorithm, *Hamiltonian Monte Carlo* (HMC), relies on theorhetical physics to compute a latent momentum variable which is applied to a hamiltonian, effectively simulating a ball rolling around the multi-dimensional sample space [@Brooks2011]. 
Despite the increased computational cost, this method is appealing because of its ability to generate proposals from distant regions of the stationary distribution with high acceptance probabilities.
This means that in practice, the HMC algorithm's stationary distribution converges on the target distribution with far fewer iterations than traditional Metropolis-Hastings implementations. 
Though HMC still requires the user to specify a step size as well as a number of steps to move the hamiltonian before considering a proposal, in practice it is much more efficient and requires fewer iterations on a model.
Additionally, a proposed extension to HMC called the *No-U-Turn Sampler* (NUTS) automatically determines the number of steps and was empirically demonstrated to perform at least equally as well as standard HMC [@Homan2014].
The approximation algorithms breifly covered in this chapter underpin modern applications for Bayesian inference, including Stan, which are discussed in the following chapter.

@@

Created at Columbia University by Bob Carpenter and Andrew Gelman, Stan is a domain specific programming language that provides tools for specifying, executing and evaluating statistical models.
Currently comprised of more than 45 core developers from a wide array of educational institutions, private entities, and research organizations, over the past 9 years the Stan project has grown to become one of the leading options for performing Bayesian inference computationally [@Stan2021].
Among the many capabilities of the Stan language, is the ability to define priors, specify dependencies between variables, and impose known limits on parameters.
After all model components have been specified, a Stan model compiles first to C++ and then to binary, which can then be executed to produce an estimation of the posterior densities of all parameters, including intermediate ones, obtained through sampling, with accompanying metadata for evaluating the sampling process.

>>>>

Imagine for a moment you are tossing a fair coin. 
There are many experiments you could perform by tossing a coin, but let us consider our quantity of interest to be the fraction of times our coin lands on a tail. 
It is clear that the number of tails, the outcome of our experiment, is dependent on the eventual realization of some random process.
A random variable $X$ is a variable whose value is dependent on the outcome(s) of a stochastic phenomenon. 
The realized value of $X$ is denoted as $x$.

@@

The r.v. $X$ is is an example of a *discrete random variable*. 
Discrete random variables can only assume discrete values. 
On the contrary, continuous random variables are useful for describing continuous sample spaces.
For example, a continuous random variable may be used to represent the outcome of an experiment measuring blossoming heights of flowers, where the data is a sequence of observations of heights at which different flowers blossomed. 
In this case, the outcome of our blossoming experiment can be any of an infinite number of real values, thus is properly modeled by a continuous random variable.

@@

In the previous example, we represented the probability of various outcomes of a coin toss experiment in a tabular format. 
Another way to represent this set of probabilities is as a *cumulative distribution function*.

@@

The cumulative distribution function $F_{X}(x)$ simply represents the probability that a random variable $X$ takes a value less than or equal to $x$ for each possible input value of $x$. 
Figure \@ref(fig:cdfimg) depicts a graphical representation of the CDF for our coin-tossing experiment.

@@

Specifically, this implies that $F'_X(x)=f_X(x)$ for all differentiable points of $F_X$. 
In plain English, this signifies that the derivative of the CDF is the PDF.

@@

Importantly while in this case, $E(X)$ corresponds well to the "peak" in the PMF, this should not be assumed to be the case unilaterally, as the same expectation would result from any distribution symmetrical about $x=3$.

@@

Conditional probability provides a way to model the probability that an event occurs, given that another event is known to have occurred.
Conditional probability, as presented in Definition \@ref(def:condprob), requires only that the event assumed to have occurred, i.e., the event we are *conditioning on*, has a nonzero probability of occurring.

@@

Conditional probability asserts that the probability of event A occurring given that event B occurs is equivalent to the probability of both A and B occurring (denoted as $A \cap B$) divided by the probability that B occurs. $P(A|B)$ is not generally equal to $P(B|A)$ 
For example, the probability that I am swimming given that I am in the water is clearly not the same as the probability that I am in the water given that I am swimming.
Example \@ref(exm:condprobe) explains how to use conditional probability to calculate the probability of rolling a 2 on a 6-sided dice, given I know the outcome of the roll is less than 4.

@@

Consider a student, Alice, who was exposed to someone with COVID-19. 
Being a responsible person, Alice decides she should get tested. 
She receives a test and the accompanying information sheet states the test is 85% accurate, meaning that 85% of the time it gives positive results to recipients who are actually positive. The sheet also says that the test yields a false positive 30% of the time, meaning that if Alice is actually negative she will still receive a positive test 30% of the time. 
The following day Alice receives a positive test.
As a student of probability, Alice recognizes that the 85% accuracy statistic only means the conditional probability that she receives a positive test given she is COVID positive ($P(+ test|covid+)$) is 85%.
However, Alice is really interested in the conditional probability that she is positive given she just tested positive, $P(covid+|test+)$.
As stated in the previous section's discussion of conditional probability, $P(A|B) \neq P(B|A)$.
Bayes' theorem, which follows intuitively from the theorem of conditional probability, provides this answer for Alice.
We can rewrite Definition \@ref(def:condprob) as 
$$P(A \cap B)=P(A|B)P(B).$$
Furthermore, it can also be stated that
$$P(B \cap A)=P(B|A)P(A)$$
Clearly, $P(A \cap B)=P(B \cap A)$ as both terms can be used interchangeably to represent the intersection of two sets A and B.
This equivalency means that we can rewrite these equations as
$$P(B|A)P(A)=P(A|B)P(B)$$
Dividing both sides of this equivalency by $P(B)$ yields Bayes' theorem, as formalized in Definition \@ref(def:bayestheorem) [@Junker2003].

@@

In the previous example, we used Bayes' theorem to evaluate the probability that Alice is COVID positive given she received a positive test. 
Bayes' theorem provided a way for Alice to apply her prior beliefs about the likelihood of actually being positive to the data she observed---a positive test. 
*Learning*, or *statistical inference*, asks a generalized version of Alice's question. 
Given some data $X_0,...,X_i \sim D$, how can we infer the distribution $D$ that generated the data $X_0,...,X_i$ [@Wasserman2004 \text{Sec} 6.1]?
Additionally, as distributions are functions of parameters $\theta_0, ..., \theta_j$, estimating these parameters given the data is one avenue for inferring $D$.
*Bayesian inference* applies Bayes' theorem to the task of statistical inference in an intuitive manner, which by definition maintains the inherent uncertainty in inference.
Contrary to the frequentist perspective, a Bayesian approach to inference expresses degrees of belief by producing posterior probability distributions for the parameters $\theta_0, ..., \theta_j$ of the distribution [@Wasserman2004 \text{Sec} 11.1]. 
Point estimates and confidence/posterior intervals can then be computed with a post hoc analysis of the posterior distribution.
Furthermore, a Bayesian approach to inference explicitly includes previous domain-specific knowledge (or lack thereof) in the learning process.
Performing Bayesian inference to compute a posterior distribution $P(\theta|X)$ for a parameter $\theta$ first requires choosing a *prior probability distribution* $P(\theta)$ with a PDF $f(\theta)$ which expresses our prior beliefs about $\theta$ before observing any data. 
In the previous example, this was Alice's previous belief about the probability of contracting COVID from any given exposure.
Expressing $P(\theta)$ as a distribution allows for the inclusion of the inherent uncertainty of Alice's preconceptions in the modeling schema and thus the computation of the posterior.
Secondly, the likelihood function $l(X|\theta)$ must be specified to express our understanding of the probability of the data given a particular value of $\theta$.
In the case of Alice's inquiry, a binomial likelihood function would be appropriate as it represents the likelihood of observing a given number of successes ($+$tests) in $0 \lor 1$ ($+ \lor -$) trials.
The final step in performing Bayesian inference is repeatedly applying Bayes' theorem to update the probability distribution of $\theta$ for each observation $X_n$ in the data $n \in (0,i)$, resulting in a posterior distribution $f(\theta|X_0,...,X_i)$.
This is done by iteratively evaluating Bayes' theorem, where the posterior for the $n-1^{th}$ iteration is the prior for the $n^{th}$ iteration.
Bayesian inference relies on Bayes' theorem for continuous variables, which utilizes density functions as presented formally in Definition \@ref(def:bayesprob).

@@

While Bayesian inference is an appealing schema, learning the parameters of real-world models using this technique requires the use of efficient algorithms to be computationally feasible.
As previously mentioned, depending on the system we may be interested in evaluating an arbitrary number of parameters $\theta_0, ..., \theta_j$ given the data, and each parameter may potentially take on an infinite number of values.
Even in the case where we only consider a small number of possible values $a$ for every $\theta_i \in (0,j)$, the evaluation of the posterior to perform a single update rapidly becomes a computationally intractable task.
This is evidenced by the runtime of the *grid approximation* or *direct discrete approximation*, which is proportional to $O(a^{j})$, where $a$ controls the granularity of the approximation [@Gelman2013 \text{Sec} 10.1].
To perform Bayesian inference on complex systems, fast algorithms are necessary to circumnavigate this curse of dimensionality.

@@

When performing Bayesian inference, fast algorithms are necessary for updating the probability distributions of parameters of interest in a computationally feasible manner.
In modeling scenarios with multiple parameters forming a multidimensional parameter space, evaluating parameters at even a relatively small number of possible values rapidly becomes intractable, as a product of the curse of dimensionality.
This problem is further exaggerated in the evaluation of hierarchical models, which are of particular interest in many application domains such as biology, economics, chemistry, and physics.
Here we discuss Markov Chain Monte Carlo (MCMC) based sampling algorithms, which allow for efficiently sampling from approximations of probability distributions.
Markov Chain Monte Carlo (MCMC) provides a methodology for sampling from the posterior curve to perform a Bayesian update. 
Simply put, if we make some sequence of draws $\theta_1,...,\theta_j$ from the posterior $p(\theta|X)$, then plotting $\theta_1,...,\theta_j$ as a histogram will provide an approximation of $p(\theta|X)$ [@Wasserman2004 \text{Sec} 11.4].
Due to the law of large numbers, this approximation can be used in place of the posterior in post-hoc analysis.
We first provide a theoretical background into Monte Carlo approximation and Markov Chains, followed by a discussion of *Metropolis-Hastings*, an algorithm for performing MCMC.

@@

Monte Carlo approximation provides a convenient method for estimating quantities of interest.
This schema will allow for drawing samples from arbitrarily complex probability distributions.
The basic example is Monte Carlo *integration* [@Wasserman2004 \text{Sec} 24.2]. 
If we want to evaluate an integral for some function $f(x)$, where 
$$I=\int_{a}^{b}f(x)dx \textrm{ ,}$$
we can approximate $I$ using Monte Carlo approximation. $f(x)$ can be alternatively expressed as two functions, $h(x)$ and $w(x)$, where $h(x)=\frac{1}{b-a}$ and $w(x)=f(x)(b-a)$ as follows.
$$I=\int_{a}^{b}f(x)dx=\int_{a}^{b}w(x)h(x)dx$$
Conveniently, $h$ is the pdf of a uniform r.v. $X$ over $[a,b]$, which means that $I$ can be rewritten in terms of expectation as
$$I=E_{f}(w(X)) \textrm{ .}$$
In conjunction with the law of large numbers, this means that if we generate a sequence of random variables from a uniform distribution $X_{0},...,X_{N}\sim unif(a,b)$ then the standard Monte Carlo integration method asserts

@@

```{example montecarl}
Suppose we have two binomially distributed r.vs---$X \sim Binom(n,p_1)$, $Y \sim Binom(m,p_2)$---and want to evaluate $\delta = p_2 - p_1$. 
This problem can be analyzed using Bayesian inference on a beta-binomial model but would require evaluating complex integrals over the joint posterior of $X$ and $Y$ to obtain the density.
Monte Carlo Integration allows for estimating an integral using repeated sampling from the two independent posterior densities---$P_1$ and $P_2$--- and plotting samples using a histogram provides a qualitative estimate for the parameter of interest, $\delta$ [@Wasserman2004].
In simplest terms, the beta prior distribution $Beta(\alpha, \beta)$ (the proper conjugate prior for the binomial likelihood) could be a flat prior $\alpha = \beta = 1$, on the parameters $p_1$ and $p_2$ [@Bloom2014].
$f(p_1,p_2|X,Y)=f(p_1|X)f(p_2|Y)$, because the two posteriors $f(p_1|X)$ and $f(p_2|Y)$ are conditionally independent given the data.
Thus the resultant posteriors are beta distributions, as follows [@Bloom2014].
$$D_1 = Beta(X+1,n-X+1)$$
$$D_2 = Beta(Y+1,m-Y+1)$$
Monte Carlo integration simulates the quantity $\delta$ by drawing $(P_1^{(1)},P_2^{(2)}),...,(P_1^{(N)},P_2^{(N)})$ from the independent beta posteriors $D_1$ and $D_2$ for $i=1,...,N$ assuming N is sufficiently large, every $\delta^{(i)}=P_2^{(i)} - P_1^{(i)}$.
Accordingly, $\delta$ is approximated using the previously established equivalency, $$\delta \approx \frac{1}{N} \sum_i^N \delta^{(i)} = \frac{1}{N} \sum_i^N (P_2^{(i)} - P_1^{(i)})$$ [@Wasserman2004].
To provide a concrete example, if the number of trials is 20, $n=m=20$, $X = 11$, and $Y = 4$ Sampling 10000 times from the in the previously described manner yields the histogram presented in Figure \@ref(fig:mon) 
```
```{r mon, fig.height=3, fig.align='center', echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Histogram displaying posterior of delta from Monte Carlo integration"}
p1 <- rbeta(10000, 12, 10)
p2 <- rbeta(10000, 5, 17)
delta <- p2 - p1
delta_df <- data.frame(delta = delta)
hist <- ggplot(delta_df, aes(x=delta)) + 
  geom_histogram(colour="black", fill="#856cb0") +
  theme_mdsr() +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.title = element_blank())
  #theme(text = element_text(size=12, 
  #                          family = "CMU Serif",
   #                         face = "plain",
   #                         colour = "black",
    #                        hjust = 0.5,
     #                       vjust = 0.5,
      #                      angle = 0,
       #                     lineheight = 0.9))

@@

*Metropolis-Hastings MCMC* is a standard algorithm for sampling from the posterior distribution to perform a Bayesian update. 
While it is not always the fastest in practice and requires manual tuning to work effectively, Metropolis-Hastings provides a foundation for more complex algorithms such as *Hamiltonian Monte Carlo* [@Brooks2011] with *No-U-Turn sampling* [@Homan2014]. 
To reiterate, the purpose of sampling in Bayesian inference is to draw from the posterior density to estimate $P(\theta|data)$ for parameters $\theta$. 
The Bayesian posterior $P(\theta|data)$, for the purposes of MCMC, is also referred to as the *target distribution*.
The basic idea of Metropolis-Hastings, to return to the previously mentioned example of approximating $I=\int h(x)f(x)dx$, is to construct a Markov chain $X_0,...,$ with a stationary distribution of $h$.
Assuming the sequence has sufficiently many observations, the law of large numbers applies and the following convergence holds.
$$\frac{1}{N}\sum_{i=0}^{N}w(X_{i})\xrightarrow{P} E(w(X))=I$$
The Metropolis-Hastings algorithm provides a method for sampling from the *stationary distribution* of a Markov chain designed in such a way that this distribution approximates the posterior distribution.
Initially, the probability of realizing a state as part of the sampled sequence is skewed, but importantly, over time the sequence will converge.
Critically, the law of large numbers guarantees that with sufficiently many iterations of Metropolis-Hastings, the generated sequence of states will converge on this stationary distribution.
Essentially, given some sequence of states $X_{0},X_{1},...,X_{i}$ from a Markov-Chain, where $X_{0}$ is chosen arbitrarily, an iteration of Metropolis-Hastings produces the next state to include in the sequence. 
As presented formally in Definition \@ref(def:methas), this is achieved by generating a *proposal* for $X_{i+1}$ from the *proposal distribution* $q(y|X_{i})$ and then accepting the proposal with some probability dependent on the relative, target probabilities of the current state $X_{i}$ and the proposal [@Wasserman2004].
```{definition, methas, name="Metropolis-Hastings MCMC"}
*$X_{i+1}$ is generated given $X_{0},X_{1},...,X_{i}$ in the following manner:*
\begin{enumerate}
  \item Sample a proposal $Y$ from the proposal distribution $Y \sim q(y|X_{i}) \textrm{ .}$
  \item Evaulate the ratio of probabilities in the stationary distribution $h(x)$ for $r(X_{i},Y)$ where $$ r(x,y)=min \{\frac{f(y)}{f(x)} \frac{q(x|y)}{q(y|x)} ,1 \} \textrm{ .}$$
  \item Compute the next state $X_{i+1}$, where $$X_{i+1}=\begin{cases}Y & \textrm{with probability }r \\ X_{i} & \textrm{with probability }1-r \end{cases} \textrm{ .}$$
\end{enumerate}
```

While Metropolis-Hastings may seem appealing because it is both memoryless and able to approximate very complex distributions, its downfall lies in the manual tuning of the *step size* parameter necessary to achieve convergence of the stationary distribution to the target distribution in a reasonable number of iterations. 
Recall that Metropolis-Hastings generates proposals by sampling from some distribution.
In the special case of *random walk Metropolis-Hastings*, this is a normal distribution centered about $X_{i}$.
Here, the standard deviation of the normal distribution, known as the step size, dictates the relative distance in the sample space between $X_{i}$ and the generated proposal.
If the step size is too low, the algorithm will make very small steps and may miss key features of the target distribution, causing the stationary distribution to require a much larger number of iterations to converge on the target distribution.
On the other hand, if the step size is too large, proposals will be overwhelmingly generated from the low-probability tails of the distribution, again resulting in a lack of convergence and poorly representative samples from the stationary distribution.
While some methods have been proposed for automatically tuning the step size parameter [@Graves2011], iterating on complex, Bayesian models to tune a parameter is not particularly efficient nor computationally feasible in many cases. 
For this reason, other approximation algorithms are used in practical implementations of Bayesian inference.
One such algorithm, *Hamiltonian Monte Carlo* (HMC), relies on theoretical physics to compute a latent momentum variable that is applied to a hamiltonian, effectively simulating a ball rolling around the multi-dimensional sample space [@Brooks2011]. 
Despite the increased computational cost, this method is appealing because of its ability to generate proposals from distant regions of the stationary distribution with high acceptance probabilities.
This means that in practice, the HMC algorithm's stationary distribution converges on the target distribution with far fewer iterations than traditional Metropolis-Hastings implementations. 
Though HMC still requires the user to specify a step size as well as a number of steps to move the hamiltonian before considering a proposal, in practice it is much more efficient and requires fewer iterations on a model.
Additionally, a proposed extension to HMC called the *No-U-Turn Sampler* (NUTS) automatically determines the number of steps and was empirically demonstrated to perform at least equally as well as standard HMC [@Homan2014].
The approximation algorithms briefly covered in this chapter underpin modern applications for Bayesian inference, including Stan, which are discussed in the following chapter.

@@

Created at Columbia University by Bob Carpenter and Andrew Gelman, Stan is a domain-specific programming language that provides tools for specifying, executing, and evaluating statistical models.
Currently comprised of more than 45 core developers from a wide array of educational institutions, private entities, and research organizations, over the past nine years the Stan project has grown to become one of the leading options for performing Bayesian inference computationally [@Stan2021].
Among the many capabilities of the Stan language, is the ability to define priors, specify dependencies between variables, and impose known limits on parameters.
After all model components have been specified, a Stan model compiles first to C++ and then to binary, which can then be executed to produce an estimation of the posterior densities of all parameters, including intermediate ones, obtained through sampling, with accompanying metadata for evaluating the sampling process.
